Estimating duration in the presence of misclassification

### 2023-07-10 cis-imperfect-testing.tex
Fix cross references

Applying the method of \cref{perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work (\eg: \autocite{cevikShedding}).

\begin{figure}
  \centering %\includegraphic{}
  \todo[inline]{Figure showing estimates without false negatives}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{perf-test} gives implausibly short estimates. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

@@

The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
Denoting the test sensitivity by $\psens$, we replace the deterministic $y_i(t)$ with a random binary variable such that:
\begin{equation}
  \prob(y_i(t) = 1) \begin{cases}
      \psens &b_i \leq t \leq e_i \\
      0 &\text{otherwise}
  \end{cases} 
\end{equation}
with $y_i(t)$ still only defined for $t \in t_i$.

However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
More plausibly, the rate of false negatives varies over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{E-ATACCC}).
\todo{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

Examination of the CIS data also provides evidence for a declining test sensitivity.
We can approximately bound the test sensitivity by assuming that all intermittent negatives are false negatives, that the negative following the last positive in an episode could be either a true or false negative, and all other negatives are true negatives.
The point estimate for the test sensitivity is the true positives / (true positives + false negatives).
In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
While the bounds are broad, they clearly suggest a declining test sensitivity over the course of an episode.
\begin{figure}
  \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
  \caption[Bounding test sensitivity using CIS data]{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
\end{figure}

I propose the following model for test sensitivity compatible with \cref{imperf-test:fig:bounding-cis-sensitivity}, consisting of a linearly declining period then a constant period:
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
\end{equation}
where $t$ is the time in days since infection
Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
\todo{Figure comparing the variable test sensitivity to fixed and CIS data}

@@

In this section, we relax the assumption of perfect testing to allow
some false negatives. We use a fairly simplified model to ensure the
likelihood is tractable. This requires modifying both $p_{ia}$ to
allow for the episode possibly being longer than observed, and
$p_{it}$ to allow for additional episodes being missed.

>>>>

Applying the method of \cref{E-perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.

\begin{figure}
  \centering %\includegraphic{}
  \todo[inline]{Figure showing estimates without false negatives}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{E-perf-test} gives implausibly short estimates. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

@@

The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
Denoting the test sensitivity by $\psens$, we replace the deterministic $y_i(t)$ with a random binary variable such that:
\begin{equation}
  \prob(y_i(t) = 1) \begin{cases}
      \psens &b_i \leq t \leq e_i \\
      0 &\text{otherwise}
  \end{cases} 
\end{equation}
with $y_i(t)$ still only defined for $t \in t_i$.
However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}

More plausibly, the rate of false negatives varies over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{E-ATACCC}).
This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

\todo[inline]{Next paragraph needs editing for clarity}
Examination of the CIS data also provides evidence for a declining test sensitivity.
We can approximately bound the test sensitivity by assuming that all intermittent negatives are false negatives, that the negative following the last positive in an episode could be either a true or false negative, and all other negatives are true negatives.
The point estimate for the test sensitivity is the true positives / (true positives + false negatives).
In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
While the bounds are broad, they clearly suggest a declining test sensitivity over the course of an episode.
\begin{figure}
  \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
  \caption[Bounding test sensitivity using CIS data]{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
\end{figure}

I propose the following model for test sensitivity compatible with \cref{imperf-test:fig:bounding-cis-sensitivity}, consisting of a linearly declining period then a constant period:
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
\end{equation}
where $t$ is the time in days since infection.
Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
\todo{Figure comparing the variable test sensitivity to fixed and CIS data}

@@

In this section, I introduce false negatives into the model of \cref{E-perf-test:sec:model}.
We use a fairly simplified model to ensure the likelihood is tractable.
This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{it}$ to allow for additional episodes being missed.
